---
title: "Practical Machine Learning Prediction Project"
date: "15/02/22"
output:
  html_document:
    theme: cerulean
---
### Background
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

### Background
The training data for this project are available here: https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv.

The test data are available here: https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv.

The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har.

### R workspace preparation
Loading libraries and set seed for reproducibility.
```{r warning=FALSE,message=FALSE}
library(ggplot2)
library(lattice)
library(caret)
library(randomForest)
set.seed(123)
```

### Load data
Loading data.
```{r}
training <- read.csv("./Input/pml-training.csv", row.names = 1)
testing <- read.csv("./Input/pml-testing.csv", row.names = 1)
```

### Data Cleansing and Preparation
We remove near zero covariates and those with more than 80% missing values. These variables will not provide necessary information for the prediction.
```{r}
zerovar <- nearZeroVar(training, saveMetrics = T)
training <- training[, !zerovar$nzv]

mval <- sapply(colnames(training), function(x) if(sum(is.na(training[, x])) > 0.8*nrow(training)){return(T)}else{return(F)})
training <- training[, !mval]
```

### Model Development
We calculate correlations between each remaining feature to the response, “classe”. We will use spearman rank based correlation because classe is a factor.

```{r}
cor <- abs(sapply(colnames(training[, -ncol(training)]), function(x) cor(as.numeric(training[, x]), as.numeric(training$classe), method = "spearman")))
summary(cor)
```

Than we will plot the two features that have highest correlation with “classe”" and color with classe to see if we can separate response based on these features.

```{r}
plot(training[, names(which.max(cor))], training[, names(which.max(cor[-which.max(cor)]))], col = training$classe, pch = 19, cex = 0.1, xlab = names(which.max(cor)), ylab = names(which.max(cor[-which.max(cor)])))
```

It appears like no strong predictors that correlate well with “classe”. Therefore the linear regression model is probably not suitable in this case. Boosting and random forests algorithms may generate more robust predictions for our data so we will use them.

### Boosting model
We will fit model with boosting algorithm and 10-fold cross validation to predict classe with all other predictors.

```{r}
boostFit <- train(classe ~ ., method = "gbm", data = training, verbose = F, trControl = trainControl(method = "cv", number = 10))
```

Than we plot accuracy of this model on the scale [0.9, 1].
```{r}
boostFit
plot(boostFit, ylim = c(0.9, 1))
```

The boosting algorithm generated a good model with accuracy = 0.997.

### Random forests model

After boosting, we fit model with random forests algorithm and 10-fold cross validation to predict “classe” with all other predictors.

```{r}
rfFit <- train(classe ~ ., method = "rf", data = training, importance = T, trControl = trainControl(method = "cv", number = 10))
rfFit
```

Then, we plot accuracy of the model on the same scale as boosting model.

```{r}
plot(rfFit, ylim = c(0.9, 1))
```

```{r}
imp <- varImp(rfFit)$importance
imp$max <- apply(imp, 1, max)
imp <- imp[order(imp$max, decreasing = T), ]
```

The random forests algorithm generated a very accurate model with accuracy close to 1. Compared to boosting model, this model generally has better performance in terms of accuracy as we see from the plots.

### Final model and prediction

When we compare the model accuracy of the two models (random forests and boosting) random forests model has overall better accuracy. The final random forests model contains 500 trees with 40 variables tried at each split. The five most important predictors in this model are r rownames(imp)[1:5]. Estimated out of sample error rate for the random forests model is 0.04% as reported by the final model.

```{r}
rfFit$finalModel
```

Predict the test set and output results for automatic grader.
```{r}
(prediction <- as.character(predict(rfFit, testing)))
```
